Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2021-12-01 manu-draft.Rmd
nelson et al. cite added
Rycyk citation, DAP processing, bloom statement moved

Recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility. At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay, particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

@@

Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).  In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a (ug/L), pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%).  Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point.  Phytoplankton samples included a mix of quantitative (cells/l) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples. Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.  Details of the sampling methods are provided in supplement.

@@

Medium bloom concentrations of the red tide species *K. brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b, low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L). The increase in *K. brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth in 2021, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread throughout lower and middle Tampa Bay. Fisheries resources were impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall of 2021.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].  Prior blooms were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions.

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the water released.  Unlike Piney Point, inorganic nitrogen concentrations of the release were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).  The fate of the phosphorus in Tampa Bay is unknown.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

>>>>

Recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility. At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack. Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.  Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) well above baseline conditions typical of surface waters in Tampa Bay. From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

@@

Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).  In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a (ug/L), pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%).  Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point.  Phytoplankton samples included a mix of quantitative (cells/l) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples. Phytoplankton blooms were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L. Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.  Details of the sampling methods are provided in supplement.

@@

Medium bloom concentrations of the red tide species *K. brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b). The increase in *K. brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth in 2021, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread throughout lower and middle Tampa Bay. Fisheries resources were impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall of 2021.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].  Prior blooms in Tampa Bay were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions. In Sarasota Bay to the south, fish activity measured by passive acoustic methods was significantly lower during a 2018 red tide event as compared to pre-bloom levels [@Rycyk20].

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the water released.  Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).  The fate of the phosphorus in Tampa Bay is unknown.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].
